Gradual Soundness: Lessons from Static Python

### 2022-01-11 pj-7.1/paper.tex
pj: fix slash error (those dang tex _s)
pj: sec 2.1.1, add carl info, cleanup, clarify

%% TODO before submission
%% - look out for PROGRESSIVE TYPES~\cite{pqk-onward-2012}, where changing annotations
%%   helps the static checker find new errors. ChkDict is one example.
%% - idea: formalism would be a SUBSET of Nom, so our burden of proof is low
%% - what are the idiomatic base SP types ... int or Int ?
%% - choose words: static vs typed, nonstatic vs dynamic vs untyped
%% - use ChkDict not ChkList ... avoid confusion
%%   ... or at least mention both, be clear the focus is Dict b/c older
%% - do the LESSONS from SP come across clearly? beware making this a "progress report"

@@

\newcommand{\spapp}[2]{#1\,(#2)}
\newcommand{\spann}[2]{#1\!:\!#2}
\newcommand{\bmname}[1]{\textbf{#1}}
\newcommand{\typefont}[1]{\mathsf{#1}}
\newcommand{\codefont}[1]{\emph{#1}}
\newcommand{\paramtype}[2]{#1[#2]}
\newcommand{\sptype}{\typefont{T}}
\newcommand{\spteval}{\typefont{S}}
\newcommand{\sptclass}{\typefont{C}}
\newcommand{\sptint}{\typefont{Int}}
\newcommand{\sptbool}{\typefont{Bool}}
\newcommand{\sptfloat}{\typefont{Float}}
\newcommand{\sptdyn}{\typefont{Dynamic}}
\newcommand{\sptobject}{\typefont{Object}}
\newcommand{\sptnone}{\typefont{None}}
\newcommand{\sptinstanceof}[1]{\paramtype{\typefont{Instance}}{#1}}
\newcommand{\sptoptional}[1]{\paramtype{\typefont{Optional}}{#1}}
\newcommand{\sptunion}[1]{\paramtype{\typefont{Union}}{#1}}
\newcommand{\sptrawpydict}{\typefont{PyDict}}
\newcommand{\sptpydict}[2]{\paramtype{\sptrawpydict}{#1, #2}}
\newcommand{\sptchkdict}[2]{\paramtype{\typefont{ChkDict}}{#1, #2}}
\newcommand{\sptenv}{\typefont{Env}}
\newcommand{\sptenvnil}{\cdot}
\newcommand{\sptenvempty}{\sptenvnil}
\newcommand{\sptvardef}[2]{\spann{#1}{#2}}
\newcommand{\sptfundef}[3]{\spapp{#1}{#2}\code{ -> }#3}
\newcommand{\sptclassdef}[4]{\mathrm{class}~\spapp{#1}{#2}:~#3;~#4}
\newcommand{\trule}[1]{\textsc{#1}}

@@

%% 2021-12-14:
%% - anything to say about Any being a "meta" type in Python typing module?
%% - 3 classes of code ... how does dynamic differ from untyped?

%% 2021-12-23: TODO
%% Carl: The behavior of dynamic type in static code should be:
%%   a) any type can flow to dynamic
%%   b) dynamic can flow to any type, but it requires the insertion of a runtime cast for soundness
%%   c) dynamic type should emit the same bytecode for any given operation that non static code would emit
%%      ... I think differing behavior around (c) is also unlikely — the structure of our
%%      code generation is that we call methods on type objects to emit byte code for
%%      various operations on a value of that type, and our dynamic type simply
%%      forwards all those calls to the default Python compiler, which should always
%%      result in dynamic type emitting the same byte code that nonstatci code would.
%%      This isn’t universal, though (i.e. for some operations, e.g. binary ops, we
%%      might explicitly handle various cases for operand type pairings, not always by
%%      calling a polymorphic method on each operand type), which could open the door
%%      to differing behavior here. But I can’t think of or find any such intentional
%%      cases, and I suspect if any did crop up we’d consider it a bug.

Like most gradual languages, \SP{} includes a dynamic type
that allows untyped expressions within a statically-typed context.
Whenever an expression or variable lacks a type annotation, \SP{} uses the
dynamic type as a default and skips most (but not all!) compile-time checks.
Every untyped Python program is thus a well-formed \SP{} program.
Furthermore, programmers can omit the annotations for specific
function arguments without losing the benefits of other type checks.
Freedom to omit annotations helps programmers convert modules to \SP{}
with little initial refactoring.

In addition to serving as a default, the dynamic type also replaces any undefined types.
For example, suppose that the type \code{X} is undefined in \SP{} and that a
program contains the annotation \code{ChkDict[int, X]}, which describes a
dictionary with integer keys and \code{X}-typed values.
\SP{} interprets this annotation as \code{ChkDict[int, dynamic]} and allows any sort
of values inside the dictionary at runtime.

At first glance, this behavior might seem odd: why not reject undefined types with an error?
The reason is to reduce friction with the Pyre annotations that are already
widely-used at Instagram.
Pyre is a mature static type system that can check more types than \SP{} knows how to enforce
efficiently, e.g., the type \code{Union[t1, ..., tn]} of arbitrary-width unions~(\cref{s:idiomatic-types}).
By replacing such types with dynamic, \SP{} can accept Pyre code that it does not yet fully understand.

%% intro sentence
The dynamic type is also different than the flexible dynamic type provided
by true gradual languages~\cite{svcb-snapl-2015}.
Replacing part of a type with dynamic can lead to both compile-time errors
and run-time errors.
In other words, \SP{} satisfies neither the static nor the dynamic gradual guarantee
for a standard type precision relation.
%% \footnote{We
%% assume a standard type precision relation. Of course, one could argue that these programs
%% do not break the gradual guarantees for a type precision relation that is tailored to \SP{}.}
%%   %% tailored = describes exactly the migrations that SP allows
\Cref{fig:gg-failure} presents two examples, one for each kind of failure:
%% \footnote{In addition, C types~(\cref{s:c-types}) are incompatible with type dynamic.}
\begin{itemize}
  \item
    \Cref{f:gg-failure-stat} presents a fully-typed class and a partially-typed subclass.
    The subclass definition raises a compile-time error because it attempts to override
    the typed \code{m1} method with another method that returns the dynamic type.

    %% 2021-12-13: TODO NOTE SWALLOW MORE
    %% I think the fundamental unusual thing we do here is our per-module
    %% incremental adoption strategy for our compiler. We don’t apply our compiler
    %% to the entire codebase, using presence or absence of type annotations as
    %% the sole indicator of whether to compile code with static or dynamic
    %% behavior. We apply our compiler only to selected opted-in modules, and the
    %% rest of the codebase is compiled with Python’s default compiler and no
    %% type-checking (it may of course be checked by Pyre but that’s unrelated to
    %% compilation.) So this clearly opens the door to differing behavior between
    %% code that is compiled by our compiler and code that isn’t, even when the
    %% former has some or many values/expressions typed as dynamic. It also makes
    %% such differences somewhat more acceptable in practice — since adoption of
    %% an individual module into our compiler is opt-in, we can swallow more
    %% behavior difference implied solely by that change.

  \item
    \Cref{f:gg-failure-dyn} presents a variant of \cref{fig:sp-example} that
    uses the type \code{ChkDict} for concrete-typed dictionaries (\cref{s:checked-type}).
    It sends a dictionary with integer values to a function that expects
    dictionaries with dynamic values.
    At runtime, the function rejects this argument because its type is not an
    exact match.
\end{itemize}
These behaviors are a consequence of \SP{} being driven by the needs of an
existing codebase.
The literature on gradual types has served only as a guide, not a mandate.

@@

    class B(A):
      def m1(self):
        return 0
    # Error: type dynamic does not match int
  \end{verbatim}
    \caption{Removing a type in \code{B} breaks the static gradual guarantee}
    \label{f:gg-failure-stat}
  \end{subfigure}
  ~
  \begin{subfigure}[t]{0.5\columnwidth}
  \begin{verbatim}
    from __static__ import ChkDict

@@

    d = ChkDict[str, int]({"A": 1})
    f(d)
    # Error: f expected ChkDict[str, dyn]
  \end{verbatim}
    \caption{Removing part of the type for \code{x} breaks the dynamic gradual guarantee}
    \label{f:gg-failure-dyn}
  \end{subfigure}
  \caption{\SP{} values performance over the gradual guarantees.}
  \label{fig:gg-failure}
\end{figure}

%% TODO carl 2021-12-24 ... very important to introduce with ChkDict ... GG not necessary only nice-to-have
%% > One area where we definitely violate the gradual guarantee and always will
%% > is primitive machine types. Unlike Python object types these do not flow to
%% > dynamic, and dynamic can never flow to them! Effectively they form a
%% > fully-typed subset of the type system that does not participate in gradual
%% > typing and can never occur in non-static code. So this violates GG because
%% > there are many examples where removing an annotation of a primitive machine
%% > type can cause static code to no longer type check.
%% > 
%% > We are ok with this violation of the gradual guarantee because the form of
%% > the gradual guarantee we really care about is not “removing an arbitrary
%% > annotation” but rather “an untyped module compiles without changes when
%% > adopted into the SP compiler” — and an untyped module cannot have
%% > pre-existing uses of the primitive machine types, since those types cannot
%% > exist outside SP.

@@

The core supported types describe basic data (integers, booleans, strings),
data structures (lists, dicts, promises),
and nominal classes.
Additional ``C types'' enable aggressive compilation of primitive
values~(\cref{s:c-types}).
Union types are well-supported statically and narrow down in accordance
with type tests (a form of occurrence typing~\cite{tf-icfp-2010,gsk-esop-2011}),
but hardly supported at run-time.
Only binary unions of a type and the none value get enforced with a run-time check
($\sptunion{\sptype, \sptnone}$, or $\sptoptional{\sptype}$).

@@

  \begin{tabular}{lll}
    %% TODO remove _
    %% TODO pick instructions
    {Instruction} & {Purpose} & {Description} \\\hline
    \bcinst{CAST} & Soundness & Assert that a value matches a type \\
    \bcinst{CHECK\_ARGS} & Soundness & Cast all inputs to a function \\[1ex]

@@

Part of this goal is met by inserting casts only in positions where the dynamic
type flows into a static type.
The materialze rule in the model illustrates this policy.
Typed functions, however, are compiled with a \bcinst{CHECK\ARGS} instruction
that casts all arguments.
By convention, this instruction always appears on the first line of a typed
function body.
The optimizer uses this convention to skip argument checks by jumping past them
when it is safe to do so; namely, whenever a typed function (or method) gets
called in a typed context.
In this way, all untyped calls get the necessary argument check and all typed
calls benefit from materialize checks at the call-site.

@@

%% TODO 2022-01-09 be careful here ... the model may handle module vars, but it's
%% just that they're untyped and (maybe) not tested

>>>>

%% TODO before submission
%% - look out for PROGRESSIVE TYPES~\cite{pqk-onward-2012}, where changing annotations
%%   helps the static checker find new errors. ChkDict is one example.
%% - choose words: static vs typed, nonstatic vs dynamic vs untyped
%% - use ChkDict not ChkList ... avoid confusion
%%   ... or at least mention both, be clear the focus is Dict b/c older
%% - do the LESSONS from SP come across clearly? beware making this a "progress report"
%% - terms: sound generic / concrete ... aren't people going to forget concrete? better keep as last resort

@@

\newcommand{\spapp}[2]{#1\,(#2)}
\newcommand{\spann}[2]{#1\!:\!#2}
\newcommand{\bmname}[1]{\textbf{#1}}
\newcommand{\typefont}[1]{\mathsf{#1}}
\newcommand{\codefont}[1]{\emph{#1}}
\newcommand{\paramtype}[2]{#1[#2]}
\newcommand{\sptype}{\typefont{T}}
\newcommand{\spteval}{\typefont{S}}
\newcommand{\sptclass}{\typefont{C}}
\newcommand{\sptX}{\typefont{X}} %% undefined type for an early example
\newcommand{\sptint}{\typefont{Int}}
\newcommand{\sptbool}{\typefont{Bool}}
\newcommand{\sptfloat}{\typefont{Float}}
\newcommand{\sptdyn}{\typefont{Dynamic}}
\newcommand{\sptobject}{\typefont{Object}}
\newcommand{\sptnone}{\typefont{None}}
\newcommand{\sptinstanceof}[1]{\paramtype{\typefont{Instance}}{#1}}
\newcommand{\sptoptional}[1]{\paramtype{\typefont{Optional}}{#1}}
\newcommand{\sptunion}[1]{\paramtype{\typefont{Union}}{#1}}
\newcommand{\sptrawpydict}{\typefont{PyDict}}
\newcommand{\sptpydict}[2]{\paramtype{\sptrawpydict}{#1, #2}}
\newcommand{\sptchkdict}[2]{\paramtype{\typefont{ChkDict}}{#1, #2}}
\newcommand{\sptenv}{\typefont{Env}}
\newcommand{\sptenvnil}{\cdot}
\newcommand{\sptenvempty}{\sptenvnil}
\newcommand{\sptvardef}[2]{\spann{#1}{#2}}
\newcommand{\sptfundef}[3]{\spapp{#1}{#2}\code{ -> }#3}
\newcommand{\sptclassdef}[4]{\mathrm{class}~\spapp{#1}{#2}:~#3;~#4}
\newcommand{\trule}[1]{\textsc{#1}}

@@

%% 2021-12-14 TODO:
%% - anything to say about Any being a "meta" type in Python typing module?
%%   2022-01-10 bg: not sure what the consequences of "meta type" are

Like most gradual languages, \SP{} includes a dynamic type
that allows untyped expressions within a statically-typed context.
Whenever an expression or variable lacks a type annotation, \SP{} uses the
dynamic type as a default and skips most \emph{but not all}\/ compile-time checks.
Every untyped Python program is thus a well-formed \SP{} program.
Furthermore, programmers can add annotations to part of a module
while leaving the rest untyped; the untyped positions receive the dynamic type.

In addition to serving as a default, the dynamic type also replaces any undefined types.
For example, suppose that the type $\sptX{}$ is undefined in \SP{} and that a
program contains the annotation $\sptchkdict{\sptint}{\sptX{}}$, which describes a
dictionary with integer keys and $\sptX{}$-typed values.
\SP{} interprets this annotation as \sptchkdict{\sptint}{\sptdyn}$ and allows any sort
of values inside the dictionary at runtime.

At first glance, this behavior might seem odd: why not reject undefined types with an error?
The reason is to reduce friction with the Pyre annotations that are already
widely-used at Instagram.
Pyre is a mature static type system that can check more types than \SP{} knows how to enforce
efficiently, e.g., the type $\sptunion{\sptype_0, \ldots, \sptype_n}$ of arbitrary-width unions~(\cref{s:idiomatic-types}).
By replacing such types with dynamic, \SP{} can accept Pyre code that it does not yet fully understand.

%% intro sentence
The dynamic type is also different than the flexible dynamic type provided
by gradual languages that satisfy the gradual guarantees~\cite{svcb-snapl-2015}.
Replacing part of a type with dynamic can lead to a compile-time error because
unlike untyped Python code, the \SP{} compiler does analyze dynamic-typed code.
Similarly, adding a dynamic type can raise a run-time error because
\SP{} does not allow interactions among different instantiations of a generic type.
%% \footnote{We
%% assume a standard type precision relation. Of course, one could argue that these programs
%% do not break the gradual guarantees for a type precision relation that is tailored to \SP{}.}
%%   %% tailored = describes exactly the migrations that SP allows
\Cref{fig:gg-failure} presents two examples, one for each kind of failure:
%% \footnote{In addition, C types~(\cref{s:c-types}) are incompatible with type dynamic.}
\begin{itemize}
  \item
    \Cref{f:gg-failure-stat} presents a fully-typed class and a partially-typed subclass.
    The subclass definition raises a compile-time error because it attempts to override
    the typed \code{m1} method with another method that returns the dynamic type.

  \item
    \Cref{f:gg-failure-dyn} presents a variant of \cref{fig:sp-example} that
    uses the type \code{ChkDict} for concrete-typed dictionaries (\cref{s:checked-type}).
    It sends a dictionary with integer values to a function that expects
    dictionaries with dynamic values.
    At runtime, the function rejects this argument because its type is not an
    exact match.
\end{itemize}
These behaviors enable efficient run-time checks and additional type-directed
optimizations.
Part of the reason \SP{} weighs these benefits more heavily than the costs
is that programmers have to opt-in to a new feature to risk the errors.
The static errors come only after enabling the \SP{} compiler on a module.
The dynamic errors come only after switching to a sound generic type.

More to the point, the \SP{} team has little interest in guarantees about
\emph{removing an arbitrary annotation}.
By contrast, there is great interest in making sure that \emph{an untyped
module compiles with minimal code changes in \SP{}}.

@@

    class B(A):
      def m1(self):
        return 0
    # Error: dynamic cannot override int
  \end{verbatim}
    \caption{SGG Violation: removing a type in the subclass \code{B} raises a static error}
    \label{f:gg-failure-stat}
  \end{subfigure}
  ~
  \begin{subfigure}[t]{0.5\columnwidth}
  \begin{verbatim}
    from __static__ import ChkDict

@@

    d = ChkDict[str, int]({"A": 1})
    f(d)
    # Error: f expected ChkDict[str, dyn]
  \end{verbatim}
    \caption{DGG Violation: removing part of the type for the parameter \code{x} raises a dynamic error}
    \label{f:gg-failure-dyn}
  \end{subfigure}
  \caption{\SP{} maintains neither the static (SGG) nor the dynamic (DGG) gradual guarantees}
  \label{fig:gg-failure}
\end{figure}

%% TODO carl 2021-12-24 ... very important to introduce machine types with ChkDict ... emphasizes that the GG is not necessary, but only nice-to-have
%% > One area where we definitely violate the gradual guarantee and always will
%% > is primitive machine types. Unlike Python object types these do not flow to
%% > dynamic, and dynamic can never flow to them! Effectively they form a
%% > fully-typed subset of the type system that does not participate in gradual
%% > typing and can never occur in non-static code. So this violates GG because
%% > there are many examples where removing an annotation of a primitive machine
%% > type can cause static code to no longer type check.

@@

The core supported types describe basic data (integers, booleans, strings),
data structures (lists, dicts, promises),
and nominal classes.
Additional ``C types'' enable aggressive compilation of primitive
values~(\cref{s:c-types}).
Union types are well-supported statically and narrow down in accordance
with type tests (a form of occurrence typing~\cite{tf-icfp-2010,gsk-esop-2011}),
but are hardly supported at run-time.
Only binary unions of a type and the none value get enforced with a run-time check
($\sptunion{\sptype, \sptnone}$, or $\sptoptional{\sptype}$).

@@

  \begin{tabular}{lll}
    {Instruction} & {Purpose} & {Description} \\\hline
    \bcinst{CAST} & Soundness & Assert that a value matches a type \\
    \bcinst{CHECK\_ARGS} & Soundness & Cast all inputs to a function \\[1ex]

@@

Part of this goal is met by inserting casts only in positions where the dynamic
type flows into a static type.
The materialze rule in the model illustrates this policy.
Typed functions, however, are compiled with a \bcinst{CHECK\_ARGS} instruction
that casts all arguments.
By convention, this instruction always appears on the first line of a typed
function body.
The optimizer uses this convention to skip argument checks by jumping past them
when it is safe to do so; namely, whenever a typed function (or method) gets
called in a typed context.
In this way, all untyped calls get the necessary argument check and all typed
calls benefit from materialize checks at the call-site.

@@

%% TODO 2022-01-09 be careful here ... the model may indeed handle module vars
%%  the thing is, they're untyped and (maybe) not tested
